### Determine the number of distinct colour classes present in the dataset.
```{r wsb-step-01}
print(" Determine the number of distinct colour classes present in the dataset.")
n_classes <- length(unique(colour_df$color))
n_classes
cat("distinct colour classes",n_classes,"\n")
```

### Fit a Quadratic Discriminant Analysis (QDA) algorithm to the classification problem.
```{r wsb-step-02}
print(" Fit a Quadratic Discriminant Analysis (QDA) algorithm to the classification problem.")
 
model_qda <- qda(color ~ r + b, data = colour_df)
model_qda
```

### Visualise the decision boundaries of the fitted QDA model in a suitable way.
```{r wsb-step-03}
print("Visualise the decision boundaries of the fitted QDA model in a suitable way.")
 
grid_points <- expand.grid(
  r = seq(0, 255, length.out = 100),
  b = seq(0, 255, length.out = 100)
)
grid_pred <- predict(model_qda, grid_points)$class
grid_points$color <- grid_pred
plot_qda_boundary <- ggplot(grid_points, aes(x = r, y = b, fill = color)) +
  geom_tile(alpha = 0.6) +
  geom_point(data = colour_df, aes(x = r, y = b, colour = color), size = 1.2, shape = 21) +
  theme_minimal() +
  labs(title = "QDA Decision Boundaries (r vs b)", x = "Red", y = "Blue")
plot_qda_boundary
```

### Test the QDA algorithm on the input (200, 0, 200).
```{r wsb-step-04}
print(" Test the QDA algorithm on the input (200, 0, 200).")
new_colour <- data.frame(r = 200, b = 200)
qda_prediction <- predict(model_qda, new_colour)
qda_prediction
```

### Determine what colour name is predicted by the algorithm for the input (200, 0, 200).
```{r wsb-step-05}
print(" Determine what colour name is predicted by the algorithm for the input (200, 0, 200).")
predicted_class <- qda_prediction$class
predicted_class
```

### (Optional) Implement k-Nearest Neighbours and compare its performance with the QDA algorithm.
```{r wsb-step-06}
print(" Learn about k-Nearest Neighbors (kNN) classification, implement it, and compare performance with QDA.")
library(class)
set.seed(82171165)
library(caret)
control <- trainControl(method = "cv", number = 10)
train_qda <- train(color ~ r + b, data = colour_df, method = "qda", trControl = control)
train_knn <- train(color ~ r + b, data = colour_df, method = "knn", tuneLength = 5, trControl = control)
qda_results <- train_qda$results
knn_results <- train_knn$results
qda_results
knn_results
```