---
title:    "STAT462 Assignment 2"
author: 
  -       "David Ewing (82171165)"
  -       "Xia Yu      (62380486)"
date:     "`r Sys.Date()`"

output:   
  pdf_document:
    latex_engine:     xelatex
    fig_caption:      true
    number_sections:  true
    toc:              true
    keep_tex:         true
    fig_crop:         false         # disable pdfcrop 
    includes:
      in_header: ../doc/fonts.tex
    
  mainfont: Arial
  fontsize: 10pt
---
```{r file-info, echo=FALSE, results='asis'}

cat("**R Markdown file name:**", knitr::current_input(), "\n\n")
cat("**R Markdown file name:**", rmarkdown::metadata$file, "\n")
cat("**Rendered on:**", format(Sys.time(), "%Y.%m.%d %H:%M:00"), "\n")
```

```{r setup, include=FALSE}
# This chunk was set up with the aid of ChatGPT.
# The intent is to load updates quietly thus not
# spending undue time with the logistics of getting 
# setup. 
setwd(rprojroot::find_root(rprojroot::is_rstudio_project))

#---------------- knitr::opts_chunk ---------------------------
#---------------- knitr::opts_chunk ---------------------------

knitr::opts_chunk$set(
  dev.args = list(
    png = list(type = "cairo")
    ),
  results = "markup",   # default("markup") "asis" "hide"	"hold"	"show"
  echo    = TRUE,      # Whether the code is shown
  eval    = TRUE,       # Whether the code is executed
  message = FALSE,      # Whether messages are printed
  warning = FALSE,     # Whether warnings are printed
  error   = FALSE       # Whether errors stop execution
  )

#---------------- loading libraries ---------------------------
#---------------- loading libraries ---------------------------

# Load configuration for reproducibility and preferred CRAN mirror
options(repos = c(CRAN = "https://cran.stat.auckland.ac.nz/"))

#library(conflicted) # before any other

# Required packages
required_packages <- c(
  "caret",         # Model training utilities
  "class",         # kNN 
  "cowplot",       # Plot composition
  "dplyr",         # Data wrangling
  "flextable",     # Summary tables
  "GGally",        # Pair plots
  "ggplot2",       # Core plotting
  "glmnet",        # Regularised regression
  "glue",          # glue
  "patchwork",      # ✅ for ggplot grid layouts
  "rsample",         # initial_split
  "kableExtra",    # Table formatting
  "knitr",         # Inline rendering
  "MASS",          # LDA/QDA and logistic
  "officer",       # Word/PDF table styling (used by flextable)
  "skimr",         # Data summaries
  "tree",
  "tidyverse",     # Core data science packages
  "tibble"         # Modern data frames
)
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

#---------------- conflict_prefer ------------------------------
#---------------- conflict_prefer ------------------------------
library(conflicted)
if ("conflicted" %in% loadedNamespaces()) {
  try(conflict_prefer("filter", "dplyr"), silent = TRUE)
  try(conflict_prefer("select", "dplyr"), silent = TRUE)
}

```
 

#---------- THIS IS UNCONDITIONAL IN PARENT ------
#---------- THIS IS UNCONDITIONAL IN PARENT ------

```{r wsb-include-helper, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)

# Only include helper if NOT running as a child
if (!isTRUE(knitr::opts_knit$get("child"))) {

  # Try to get current Rmd file path
  this_file <- knitr::current_input()
  if (is.null(this_file) && rstudioapi::isAvailable()) {
    this_file <- rstudioapi::getSourceEditorContext()$path
  }

  if (!is.null(this_file) && file.exists(this_file)) {
    project_subdir <- dirname(normalizePath(this_file))

    helper_path <- file.path(project_subdir, "HELPER.R")
    cat(knitr::knit_child(helper_path, envir = knitr::knit_global()))
  } else {
    cat("Unable to detect Rmd path.\n")
  }
}



```



 

 







```{r test}

this_file <- knitr::current_input()

# Fallback if knitting doesn't give you the file
if (is.null(this_file) && rstudioapi::isAvailable()) {
  this_file <- rstudioapi::getSourceEditorContext()$path
}

# Only proceed if we now have a valid path
if (!is.null(this_file) && file.exists(this_file)) {
  this_path <- normalizePath(this_file)
  this_dir  <- dirname(this_path)
  cat("Rmd is in folder:", this_dir)
} else {
  cat("Unable to detect Rmd path.\n")
}



 
```


```{r wsb-step-03-split}
cat("### Split the dataset into a training set (80%) and a test set (20%)
")

# heart_df exists at run time

h_vars <- c("DEATH", "GLUCOSE", "SYSBP", "AGE")
h_clean <- heart_df |>
  select(all_of(h_vars)) |>
  drop_na() |>
  mutate(DEATH = as.factor(DEATH))

# Stratified split : DEATH
set.seed(82171165)
h_split   <- initial_split(h_clean, prop = 0.8, strata = DEATH)
h_train   <- training(h_split)
h_test    <- testing(h_split)

# For visualisation only: relabel  "LIVE" and "DEAD"
levels = c(0, 1)
labels = c("LIVE", "DEAD")


hp_clean <- h_clean |>
  mutate(DEATH = factor(DEATH, levels = levels, labels = labels))
hp_train <- h_train |>
  mutate(DEATH = factor(DEATH, levels = levels, labels = labels))
hp_test <- h_test |>
  mutate(DEATH = factor(DEATH, levels = levels, labels = labels))

h_actual_class  <- h_test$DEATH        # true observed values
hp_actual_class <- hp_test$DEATH

```
### Visualise the relationship between DEATH, GLUCOSE and SYSBP

<!-- Inserted from WBS: heart-wsb-step-02 -->
```{r wsb-step-04-visual}
cat("### Visualise the relationship between DEATH, GLUCOSE and SYSBP\n")

ggp_relationship <- ggplot(hp_clean, aes(x = GLUCOSE, y = SYSBP, colour = DEATH)) +
  geom_point(alpha = 0.6) +
  labs(title = "DEATH vs GLUCOSE and SYSBP", x = "GLUCOSE", y = "SYSBP")

# 1. Distribution of SYSBP
ggp_sysbp <- ggplot(hp_test, aes(x = SYSBP)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of SYSBP", x = "SYSBP", y = "Count")

# 2. Distribution of GLUCOSE
ggp_glucose <- ggplot(hp_test, aes(x = GLUCOSE)) +
  geom_histogram(bins = 40, fill = "darkorange", colour = "white") +
  labs(title = "Distribution of GLUCOSE", x = "GLUCOSE", y = "Count")

# 3–6. Combination of linear/log axis scales
base <- ggplot(hp_test, aes(x = GLUCOSE, y = SYSBP)) +
  geom_point(alpha = 0.6) +
  labs(title = NULL, x = "GLUCOSE", y = "SYSBP")

p_lin_lin <- base + ggtitle("Linear X / Linear Y")
p_log_lin <- base + scale_x_log10() + ggtitle("Log X / Linear Y")
p_lin_log <- base + scale_y_log10() + ggtitle("Linear X / Log Y")
p_log_log <- base + scale_x_log10() + scale_y_log10() + ggtitle("Log X / Log Y")

# Combine 4 into a grid
p_combined <- (p_lin_lin | p_log_lin) / (p_lin_log | p_log_log)

# Output all plots
print(ggp_relationship)
print(ggp_sysbp)
print(ggp_glucose)
print(p_combined)

pretty_ggplot(ggp_relationship, title = "Relationship: DEATH vs GLUCOSE and SYSBP")
pretty_ggplot(ggp_sysbp, title = "Distribution of SYSBP")
pretty_ggplot(ggp_glucose, title = "Distribution of GLUCOSE")
pretty_ggplot(p_lin_lin, title = "Linear X / Linear Y")
pretty_ggplot(p_log_lin, title = "Log X / Linear Y")
pretty_ggplot(p_lin_log, title = "Linear X / Log Y")
pretty_ggplot(p_log_log, title = "Log X / Log Y")
pretty_ggplot(p_combined, title = "Log/Linear Scale Comparison Grid")

```

### Form an initial hypothesis of what to look for when doing the classification


<!-- Inserted from WBS: heart-wsb-step-03 -->
```{r  heart-wsb-step-03}
print("We hypothesise that high GLUCOSE and high SYSBP are associated with higher DEATH risk.")
```

### Fit a logistic regression model on the training set


<!-- Inserted from WBS: heart-wsb-step-04 -->
```{r  heart-wsb-step-04}
hp_glm_metrx <- glm(DEATH ~ GLUCOSE + SYSBP + AGE, data = hp_train, family = "binomial")
pretty_glm(hp_glm_metrx)
```


```{r wsb-step-05-model}
# Fit a logistic regression model on training data
# Summarize using pretty_glm
hp_model <- glm(DEATH ~ GLUCOSE + SYSBP + AGE, data = hp_train, family = "binomial")
pretty_glm(hp_model)
```





### Compute the misclassification rate on the test set

```{r wsb-step-06}
cat("### Compute the misclassification rate on the test set\n")

# Generate predicted probabilities and classes
hp_pred_probs <- predict(hp_model, newdata = hp_test, type = "response")
hp_pred_class <- predict_classes(hp_pred_probs, threshold = 0.5)

# Compute actual class labels
hp_actual_class <- factor(hp_test$DEATH, levels = c("LIVE", "DEAD"))

# Calculate confusion matrix
hp_cmtx <- make_conf_matrix(hp_pred_class, hp_actual_class)

# Evaluate confusion matrix
hp_metrics <- eval_conf_matrix(hp_cmtx)
hp_mtrx_df <- summarise_metrics(hp_metrics)

# Convert confusion matrix to printable data frame
hp_cmtx_df <- as.data.frame.matrix(hp_cmtx)
hp_cmtx_df <- cbind(Predicted = rownames(hp_cmtx_df), hp_cmtx_df)
rownames(hp_cmtx_df) <- NULL

# Print formatted summaries
print(pretty_df(hp_cmtx_df, title = "Logistic Regression Confusion Matrix (Threshold = 0.5)"))
print(pretty_df(hp_mtrx_df, title = "Logistic Regression Performance (Threshold = 0.5)"))

 
```{r wsb-step-06-predict}
# Generate predicted probabilities and classify using predict_classes
hp_pred_probs <- predict(hp_model, newdata = hp_test, type = "response")
hp_pred_class <- predict_classes(hp_pred_probs, threshold = 0.5)
```

### Compute the confusion matrix on the test set
```{r wsb-step-07-eval-cm}
cat("### Logistic Regression Evaluation (Threshold = 0.50)\n")



hp_pred_probs <- predict(hp_glm_metrx, newdata = hp_test, type = "response")
hp_pred_class <- predict_classes(hp_pred_probs, threshold = 0.5)
hp_actual_class <- factor(hp_test$DEATH, levels = c("LIVE", "DEAD"))

hp_cmtx_test <- make_conf_matrix(hp_pred_class, hp_actual_class)
hp_metrics <- eval_conf_matrix(hp_cmtx_test)
hp_mtrx_df <- summarise_metrics(hp_metrics)

hp_cmtx_df <- as.data.frame.matrix(hp_cmtx_test)
hp_cmtx_df <- cbind(Predicted = rownames(hp_cmtx_df), hp_cmtx_df)
rownames(hp_cmtx_df) <- NULL

print(pretty_df(hp_cmtx_df, title = "Confusion Matrix (Threshold = 0.50)"))
print(pretty_df(hp_mtrx_df, title = "Logistic Regression Performance (Threshold = 0.50)"))
```
#  . Predict on test set and evaluate
```{r wsb-step-08-visualise}
# Predict using QDA 
# Add predicted class to test data
# Define custom confusion matrix function
# Create the confusion matrix
c_qda_pred <- predict(c_qda_train, newdata = c_test)
c_test$.pred_class <- c_qda_pred$class
c_conf_mat <- make_conf_matrix(c_test$.pred_class, c_test$colour)

# View it
print(c_conf_mat)


<!-- Inserted from WBS: heart-wsb-step-07 -->
```{r  heart-wsb-step-07}

h_ggp_1 = ggplot(hp_test, aes(x = GLUCOSE, y = SYSBP, colour = hp_pred_class)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Logistic Regression Decision Boundary (approx)",
    colour = "Predicted"
  )

```

### Comment on the goodness of fit
```{r  heart-wsb-step-08}
#refere to hp_mtrx_df 
```

### Supporting numeric metrics for evaluation criteria


<!-- Inserted from WBS: heart-wsb-step-09 -->
```{r  heart-wsb-step-09}
hp_pred_class_10 <- ifelse(hp_pred_probs > 0.1, 1, 0)
hp_pred_class_10 <- factor(hp_pred_class_10, levels = c(0, 1), labels = c("LIVE", "DEAD"))
```

### Recompute misclassification, confusion, and visualisation for 10% threshold
# Confusion Matrix (Threshold = 0.10)


<!-- Inserted from WBS: heart-wsb-step-10 -->
```{r wsb-step-10}
#hp_misclass_rate_10 <- mean(hp_pred_class_10 != hp_actual_class)
#hp_cmtx_10 = table(Predicted = h_pred_class_10, Actual = h_actual_class)
```

### Compare logistic regression and discriminant analysis
```{r heart-wsb-step-10.1}
cat("### Logistic Regression Evaluation (Threshold = 0.10)\n")

hp_pred_probs <- predict(hp_glm_metrx, newdata = hp_test, type = "response")
hp_pred_class_10 <- predict_classes(hp_pred_probs, threshold = 0.10)
hp_actual_class <- factor(hp_test$DEATH, levels = c("LIVE", "DEAD"))

hp_cmtx_test_10 <- make_conf_matrix(hp_pred_class_10, hp_actual_class)
hp_metrics_10 <- eval_conf_matrix(hp_cmtx_test_10)
hp_mtrx_df_10 <- summarise_metrics(hp_metrics_10)

hp_cmtx_df_10 <- as.data.frame.matrix(hp_cmtx_test_10)
hp_cmtx_df_10 <- cbind(Predicted = rownames(hp_cmtx_df_10), hp_cmtx_df_10)
rownames(hp_cmtx_df_10) <- NULL

print(pretty_df(hp_cmtx_df_10, title = "Confusion Matrix (Threshold = 0.10)"))
print(pretty_df(hp_mtrx_df_10, title = "Logistic Regression Performance (Threshold = 0.10)"))
```

### QDA Model
<!-- Inserted from WBS: heart-wsb-step-11 -->
```{r heart-wsb-step-11}
hp_qda_model <- qda(DEATH ~ GLUCOSE + SYSBP + AGE, data = hp_train)```
```


<!-- Inserted from WBS: heart-wsb-step-11 -->
### QDA Evaluation
<!-- Inserted from WBS: heart-wsb-step-11.1 -->
```{r heart-wsb-step-11.1}
cat("### QDA Evaluation\n")

# Build QDA model using labeled data
hp_qda_model <- qda(DEATH ~ GLUCOSE + SYSBP + AGE, data = hp_train)

# Predict posterior probabilities and classes
qda_output <- predict(hp_qda_model, newdata = hp_test)
hp_qda_pred_probs <- qda_output$posterior[, "DEAD"]
hp_qda_pred_class <- predict_classes(hp_qda_pred_probs, threshold = 0.5)

# Define actual classes
hp_actual_class <- factor(hp_test$DEATH, levels = c("LIVE", "DEAD"))

# Confusion matrix and metrics
hp_cmtx_qda <- make_conf_matrix(hp_qda_pred_class, hp_actual_class)
hp_metrics_qda <- eval_conf_matrix(hp_cmtx_qda)
hp_mtrx_df_qda <- summarise_metrics(hp_metrics_qda)

# Format confusion matrix for display
hp_cmtx_df_qda <- as.data.frame.matrix(hp_cmtx_qda)
hp_cmtx_df_qda <- cbind(Predicted = rownames(hp_cmtx_df_qda), hp_cmtx_df_qda)
rownames(hp_cmtx_df_qda) <- NULL

# Print results
print(pretty_df(hp_cmtx_df_qda, title = "QDA Confusion Matrix"))
print(pretty_df(hp_mtrx_df_qda, title = "QDA Model Performance"))

```
 

### Visualise the decision boundaries of the fitted QDA model in a suitable way

<!-- Inserted from WBS: heart-wsb-step-12 -->
```{r heart-wsb-step-12}
print("Visualise the decision boundaries of the fitted QDA model in a suitable way.")

# A simple visualisation by predicted class
hp_qda_probs <- predict(hp_qda_model, newdata = hp_test)$posterior[, 2]
hp_qda_pred_class <- predict_classes(hp_qda_probs, threshold = 0.5)


ggp_qda_boundary <- ggplot(hp_test, aes(x = GLUCOSE, y = SYSBP, colour = hp_qda_pred_class)) +
  geom_point(alpha = 0.5) +
  labs(title = "QDA Decision Boundary (approx)", colour = "Predicted")



print(ggp_qda_boundary)
```

### Identify strong risk factors and communicate results


<!-- Inserted from WBS: heart-wsb-step-13 -->
### Fit a tree-based model to the training set
<!-- Inserted from WBS: heart-wsb-step-13 -->
```{r heart-wsb-step-13}
cat("### Fit a Tree-Based Model to the Training Set\n")

# Fit classification tree to hpl_train
hp_tree_model <- tree(DEATH ~ GLUCOSE + SYSBP + AGE, data = hp_train)

# Optional: Display summary of the model
print(summary(hpl_tree_model))

# Optional: Visualise the tree
plot(hpl_tree_model)
text(hpl_tree_model, pretty = 0)

```

### Tree-Based or Optional Model Evaluation
```{r heart-wsb-step-13.1}
cat("### Tree-Based or Optional Model Evaluation\n")

# Generate predicted probabilities and classes
hp_pred_probs <- predict(hp_tree_model, newdata = hp_test, type = "response")
hp_pred_class <- predict_classes(hp_pred_probs[, "DEAD"], threshold = 0.5)
hp_actual_class <- factor(hp_test$DEATH, levels = c("LIVE", "DEAD"))

# Build confusion matrix
hp_cmtx <- make_conf_matrix(hp_pred_class, hp_actual_class)
hp_metrics <- eval_conf_matrix(hp_cmtx)
hp_mtrx_df <- summarise_metrics(hp_metrics)

# Format for display
hp_cmtx_df <- as.data.frame.matrix(hp_cmtx)
hp_cmtx_df <- cbind(Predicted = rownames(hp_cmtx_df), hp_cmtx_df)
rownames(hp_cmtx_df) <- NULL

# Print both confusion matrix and metrics
print(pretty_df(hp_cmtx_df, title = "Tree-Based Model Confusion Matrix"))
print(pretty_df(hp_mtrx_df, title = "Tree-Based Model Performance"))

```


```
