---
title: 
  - "STAT462 Assignment 2 - Classification"
author: 
  - Xia Yu (62380486)
  - David Ewing (82171165)
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    df_print: paged
---

```{r, include = F, echo = F, eval=T}
# Environment setup: Install and load required packages
options(repos = c(CRAN = "https://cran.stat.auckland.ac.nz/"))
required_packages <- c("conflicted", "ggplot2", "dplyr", "plotly", "tidyr", "caret", "knitr", "reshape2","readr","tidyverse","skimr","pROC","MASS")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}

library(conflicted)
conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  tidyr::expand,
  plotly::layout,
  pROC::auc
)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, width = 70, cache = FALSE)
knitr::opts_knit$set(root.dir = getwd())
```

# Appendix - Common Functions

### Function: read_csv_from_zip

Unzips a specified ZIP file to read a CSV file while keeping specified columns.

Usage: read_csv_from_zip(zip_filepath, csv_filename, columns_to_keep)

```{r, include=T, eval=T}
read_csv_from_zip <- function(zip_filepath, csv_filename, columns_to_keep) {
  unzip(zip_filepath, files = csv_filename) %>%
    read_csv() %>%
    select(columns_to_keep)
}
```

### Function get_csv_column_names_from_zip

Extracts column names from a CSV file within a ZIP archive without reading the entire file.

Usage: get_csv_column_names_from_zip(zip_filepath, csv_filename)

```{r, include=T, eval=T}
get_csv_column_names_from_zip <- function(zip_filepath, csv_filename) {
  # 1. Extract the CSV file to a temporary directory
  unzip(zip_filepath, files = csv_filename, exdir = tempdir())
  temp_csv_path <- file.path(tempdir(), csv_filename)

  # 2. Read the CSV file, reading only the header row (n_max = 0)
  data <- read_csv(temp_csv_path, n_max = 0)

  # 3. Get the column names
  column_names <- names(data)

  # 4. Return the column names
  return(column_names)
}
```

### Function: split_data

Splits a dataset into training and test sets based on a specified training ratio.

Usage: split_data(data, train_ratio = 0.8)

```{r, include=T, eval=T}
# Function to split data into training and test sets
split_data <- function(data, train_ratio = 0.8) {
  # Set a seed for reproducibility and to minimize RAM usage
  set.seed(62380486)
  # set.seed(2)
  # validate train_ratio range
  if (train_ratio <= 0 || train_ratio >= 1) {
  stop("Error: train_ratio must be between 0 and 1 (exclusive).")
}
  # Randomly select the specified percentage of indices for the training set
  train_ind <- sample(1:nrow(data), 
                      size = floor(train_ratio * nrow(data)),
                      replace = FALSE)
  
  # Use the remaining indices for the test set
  test_ind <- setdiff(1:nrow(data), train_ind)
  
  # Create training data using the selected indices
  train_data <- data[train_ind, , drop = FALSE]
  rownames(train_data) <- NULL

  # Create test data using the remaining indices
  test_data <- data[test_ind, , drop = FALSE]
  rownames(test_data) <- NULL
  
  # Return both training and test data as a list
  return(list(train = train_data, test = test_data))
}

```

### Function: confusion_matrix_cal

Calculates the confusion matrix and misclassification rate for a classification model.

Usage: confusion_matrix_cal(model = logreg.fit, test_data = sp_data\$test, threshold = 0.1, outcome_variable = "DEATH")

```{r, include=T, eval=T}
confusion_matrix_cal <- function(model=logreg.fit, test_data=sp_data$test, threshold = 0.1, outcome_variable = "DEATH") {
  # 1. Predict probabilities (using the test set)
  predicted_probabilities <- predict(model, newdata = test_data, type = "response")

  # 3. Convert to binary classification
  predicted_classes <- ifelse(predicted_probabilities > threshold, 1, 0)

  # 4. Calculate the confusion matrix (using the test set)
  confusion_matrix <- table(Actual = test_data[[outcome_variable]], Predicted = predicted_classes)
  # print(paste("------"))
  # print(paste("threshold:",threshold))
  # print(confusion_matrix)
  
  # 5. Calculate the misclassification rate
  misclassification_rate <- (confusion_matrix[1, 2] + confusion_matrix[2, 1]) / sum(confusion_matrix)
  # print(paste("misclassification_rate:",misclassification_rate))
  
  return(list(misclassification_rate = misclassification_rate, confusion_matrix = confusion_matrix))
 }

```

### Function: roc_curve_plot

Generates and plots the ROC curve for a classification model, and calculates the AUC.

Usage: roc_curve_plot(model = logreg.fit, test_data = sp_data\$test, outcome_variable = "DEATH")

```{r, include=T, eval=T}
# ROC curve function
roc_curve_plot <- function(model = logreg.fit, test_data = sp_data$test, outcome_variable = "DEATH") {
  # 1. Predict probabilities
  predicted_probabilities <- predict(model, newdata = test_data, type = "response")

  # 2. Create ROC object, force direction to make x-axis consistent
  roc_obj <- roc(test_data[[outcome_variable]], predicted_probabilities, direction = "<")

  # 3. Plot ROC
  plot(roc_obj,
       main = "ROC Curve",
       col = "blue",
       lwd = 2,
       xlab = "1 - Specificity (False Positive Rate)",
       ylab = "Sensitivity (True Positive Rate)",
       print.auc = TRUE,
       auc.polygon = TRUE,
       auc.polygon.col = "skyblue2")

  return(roc_obj)
}


```

### Function: roc_qda_curve_plot

Generates and plots the ROC curve specifically for a QDA classification model, and calculates the AUC.

Usage: roc_qda_curve_plot(model, test_data, outcome_variable = "DEATH")

```{r}
# Custom Function for Plotting ROC Curve of QDA Model
roc_qda_curve_plot <- function(model, test_data, outcome_variable = "DEATH") {
  # 1. Get Predicted Posterior Probabilities
  predicted_probabilities <- predict(model, newdata = test_data)$posterior[, 2]  # Probability for Class 1

  # 2. Create ROC Object (force consistent direction)
  roc_obj <- roc(test_data[[outcome_variable]], predicted_probabilities, direction = "<")

  # 3. Plot ROC Curve
  plot(roc_obj,
       main = "QDA Model - ROC Curve",
       col = "darkgreen",
       lwd = 2,
       xlab = "1 - Specificity (False Positive Rate)",
       ylab = "Sensitivity (True Positive Rate)",
       print.auc = TRUE,
       auc.polygon = TRUE,
       auc.polygon.col = "lightgreen"
  )

  return(roc_obj)
}

```

\newpage

# Question A: Applying Logistic Regression to predict mortality from blood glucose and blood pressure

```{r load-Q1, child="./Q1.Rmd", eval=T}

```

\newpage

# Question B: Predicting color name from RGB values, using discriminant analysis

```{r load-Q2, child="./Q2.Rmd", eval=T}

```
