---
title:    "STAT462 Assignment 2"
author: 
  -       "David Ewing (82171165)"
  -       "Xia Yu      (62380486)"
date:     "`r Sys.Date()`"
output:   
  pdf_document:
    latex_engine: xelatex
    fig_caption:     true
    number_sections: true
    toc:             true
    keep_tex:        true
    fig_crop:        false         # disable pdfcrop 
    includes:
      in_header: ../doc/fonts.tex
fontsize: 11pt
geometry: margin=1in
---
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    
mainfont: Arial
fontsize: 10pt

```{r setup, include=FALSE}
# This chunk was set up with the aid of ChatGPT.
# The intent is to load updates quietly thus not
# spending undue time with the logistics of getting 
# setup. 

#---------------- loading libraries ---------------------------
#---------------- loading libraries ---------------------------
# Load configuration for reproducibility and preferred CRAN mirror
options(repos = c(CRAN = "https://cran.stat.auckland.ac.nz/"))

library(conflicted) # befor any other

# Required packages
required_packages <- c(
  "caret",         # Model training utilities
  "class",         # kNN 
  "cowplot",       # Plot composition
  "dplyr",         # Data wrangling
  "flextable",     # Summary tables
  "GGally",        # Pair plots
  "ggplot2",       # Core plotting
  "glmnet",        # Regularised regression
  "kableExtra",    # Table formatting
  "knitr",         # Inline rendering
  "MASS",          # LDA/QDA and logistic
  "officer",       # Word/PDF table styling (used by flextable)
  "skimr",         # Data summaries
  "tidyverse",     # Core data science packages
  "tibble"         # Modern data frames
)
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

#---------------- conflict_prefer ------------------------------
#---------------- conflict_prefer ------------------------------
conflict_prefer("filter", "dplyr"); conflict_prefer("select", "dplyr")

set.seed(82171165)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))


# ---- Pretty Summary  with Type Summary and Preview ------
# ---- Pretty Summary  with Type Summary and Preview ------
# works with flextable; provides a summary of column types
# and a preview

pretty_summary <- \(df) {   # works with flextable 
  select(df, where(is.numeric)) |>
    summarise(
      across(
        everything(),
        .fns = list(
          Mean     = \(x) mean(x, na.rm = TRUE),
          Median   = \(x) median(x, na.rm = TRUE),
          Min      = \(x) min(x, na.rm = TRUE),
          Max      = \(x) max(x, na.rm = TRUE),
          IQR      = \(x) IQR(x, na.rm = TRUE),
          nNA      = \(x) sum(is.na(x))
        )
      )
    ) |>
    pivot_longer(
      cols = everything(),
      names_to = c("Variable", ".value"),
      names_sep = "_"
    ) |>
    mutate(across(where(is.numeric), round, 2)) |>
    flextable() |>
    set_header_labels(
      Variable = "Feature",
      Mean     = "Mean",
      Median   = "Median",
      Min      = "Min",
      Max      = "Max",
      IQR      = "Interquartile Range",
      nNA      = "Missing Values"
    ) |>
    autofit() |>
    theme_vanilla()
}

# ---- Pretty csv_read with Type Summary and Preview ------
# ---- Pretty csv_read with Type Summary and Preview ------ 
# works with flextable; provides a summary of column types
# and a preview

pretty_read_csv <- \(path, n = 0) {                   # to work with flextable
  df <- readr::read_csv(path, show_col_types = FALSE)
  types <- purrr::map_chr(df, typeof)
  type_df <- data.frame(Column = names(df), Type = types, stringsAsFactors = FALSE)

  print(flextable::flextable(type_df) |>
          flextable::autofit() |>
          flextable::set_caption(caption = "Column Types in Loaded CSV"))

  if (n > 0) print(head(df, n))
  return(df)
}



pretty_df <- function(df, title = NULL, fontsize = 10, max_rows = 20) {
  df_show <- head(df, max_rows)  # Limit to first `max_rows` rows for readability
  
  ft <- flextable(df_show) |>
    fontsize(size = fontsize, part = "all") |>
    autofit() |>
    theme_booktabs() |>
    align(align = "center", part = "all") |>
    padding(padding = 5, part = "all") 
  
  if (!is.null(title)) {
    ft <- set_caption(ft, caption = title)
  }
  
  return(ft)
}

 

pretty_ggplot <- function(plot, title = "ggplot Summary") {
  if (!inherits(plot, "gg")) stop("Input must be a ggplot object.")

  # Extract key components
  plot_data <- tryCatch(plot$data, error = function(e) NULL)
  geoms <- sapply(plot$layers, function(layer) class(layer$geom)[1])
  mappings <- plot$mapping
  
  # Pull out global aesthetics as strings
  global_aes <- names(mappings)
  global_vals <- sapply(mappings, function(x) rlang::expr_text(x))

  # Additional metadata
  plot_title <- plot$labels$title %||% ""
  x_lab <- plot$labels$x %||% ""
  y_lab <- plot$labels$y %||% ""
  colour_lab <- plot$labels$colour %||% plot$labels$color %||% ""

  # Build a metadata data frame
  df <- data.frame(
    Component = c("Title", "X Axis", "Y Axis", "Colour Legend", "Geoms", global_aes),
    Value = c(plot_title, x_lab, y_lab, colour_lab, paste(geoms, collapse = ", "), global_vals),
    stringsAsFactors = FALSE
  )

  # Format with flextable
  flextable(df) |>
    set_caption(caption = title) |>
    autofit() |>
    theme_booktabs() |>
    fontsize(size = 10, part = "all") |>
    align(align = "left", part = "all") |>
    padding(padding = 4)
}

 

pretty_model <- function(model, title = "Model Summary", digits = 3) {
  if (!inherits(model, "glm")) stop("Only 'glm' objects are currently supported.")
  
  # Use broom::tidy() for clean coefficient table
  tidy_tbl <- broom::tidy(model) |>
    mutate(across(where(is.numeric), ~ round(.x, digits)))

  # Optionally include model fit stats (like AIC, null deviance, etc.)
  stats_tbl <- data.frame(
    Metric = c("Formula", "AIC", "Null deviance", "Residual deviance"),
    Value = c(
      as.character(formula(model)),
      round(AIC(model), digits),
      round(model$null.deviance, digits),
      round(model$deviance, digits)
    ),
    stringsAsFactors = FALSE
  )
  
  # Format coefficient table
  ft_coef <- flextable(tidy_tbl) |>
    set_caption(caption = title) |>
    autofit() |>
    theme_booktabs() |>
    fontsize(size = 10, part = "all") |>
    align(align = "center", part = "all") |>
    padding(padding = 4, part = "all")

  # Format stats table separately (optional)
  ft_stats <- flextable(stats_tbl) |>
    set_caption(caption = paste(title, "(Fit Stats)")) |>
    autofit() |>
    theme_booktabs() |>
    fontsize(size = 9, part = "all") |>
    align(align = "left", part = "all") |>
    padding(padding = 4, part = "all")
  
  # Return both as a list for separate rendering
  list(Coefficients = ft_coef, Stats = ft_stats)
}



```

# Introduction

This report addresses two distinct classification tasks:

- **Question A** involves predicting mortality (`DEATH`) using physiological variables from `heart.csv`.
- **Question B** focuses on classifying colour names from RGB values in `colour_df_train.csv`.

Each question is handled independently after common data preparation.




# Data Preparation

```{r load-data}
# Load heart and colour datasets from zip

# Set base directory
# Assign named paths based on filename match
# load the dataframes via pretty_read_csv 
# preview structure
unzip_dir <- "../data/unzipped"
zip_path  <- "../data/data_assignment_2.zip"
csv_files <- unzip(zip_path, list = TRUE)$Name  # Extract filenames from the zip
target_paths <- file.path(unzip_dir, csv_files)

index_color <- grep("color", csv_files)
index_heart <- grep("heart", csv_files)
path_colour <- target_paths[index_color]
path_heart  <- target_paths[index_heart]
 
heart_df  <- pretty_read_csv(path_heart)
colour_df <- read_csv(path_colour, col_names = TRUE)
str(heart_df)
str(colour_df)
```
.


# Question A: Predicting Mortality (`heart.csv`)

```{r child="../doc/heart-analysis.Rmd"}
```



## A(a) Train-Test Split

## A(b) Exploratory Visualisation

## A(c) Logistic Regression

- (i) Misclassification Rate  
- (ii) Confusion Matrix  
- (iii) Decision Boundary Visualisation

## A(d) Sensitivity-Prioritised Model

## A(e) Comparison with Discriminant Analysis

## A(f) Feature Importance and Risk Factors




# Question B: Colour Classification (`colour_df_train.csv`)

#```{r load-color-analysis}
# Source all relevant functions and logic for Question B
#source("colour-analysis-B1-B11-CHUNKED.Rmd", echo = TRUE, local = knitr::knit_global())
#```


```{r child="../doc/colour-analysis.Rmd"}
```

## B(a) Number of Classes

## B(b) QDA Classification and Visualisation

## B(c) Classification of Colour (200, 0, 200)

## B(d) kNN Comparison (Optional)




# Discussion

This section synthesises findings from both classification tasks, comparing model performance and interpretability.

# Conclusion

Summary of key results, limitations, and future recommendations.



